\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}


%\pagestyle{empty}

\setlength{\parskip}{1em}

\setlength{\parindent}{0em}


\begin{document}
\begin{flushright}
Zak Webb
\hfill
University of Waterloo\\
Institute for Quantum Computing\\
200 University Ave W\\
Waterloo, ON, Canada, N2L 3G1\\
 zakwwebb@gmail.com\\
~
\\
\today
\end{flushright}

To whom it may concern:

I am applying for a Doctoral Thesis Completion award.  I currently planning to defend, and submit, my thesis in September 2016, so that I will not be an enrolled student for the fall term.  During this term, I am planning to only write my thesis and prepare my defense, and in particular I will not hold any employment during this term.  I also do not hold any external awards for the spring term.

This award will be of extreme help to me.  Two years ago (October 2014), my advisor moved universities, and is now located at the University of Maryland.  As I was originally planning to defend my thesis by September 2015, I planned to stay here at the University of Waterloo, to work on my thesis and communicate with my advisor electronically.  However, I ran into some mental health issues that eventually led to my removal from the physics program.  I have since sought help, but this has resulted in my thesis taking longer to write than originally planned.  In particular, my advisor's Canadian funding has run out, causing my financial support to become rather difficult.  (A related problem has resulted in my funding for the fall term only recently being paid via GRS). 

Essentially, at this point my thesis is nearing completion, with extensions and generalizations for several of my previous results.  However, I still need to turn the collection of new results into a coherent document, which will take some time.   Having secured funding for this term will greatly ease my mind, and allow me to focus on finishing the document.



Sincerely,

\vspace{1in}

{\bfseries Zak Webb}\\
%
\vfill%
{\slshape Attached: curriculum vit\ae{}, unofficial transcript, latest degree progress report, doctoral thesis completion award application}
\end{document}